Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-01-20 paper.qmd
Explain why PaCMAP doesn't preserve local structure

The models are fitted to these layouts, but we focus on a single cluster to illustrate the curious detail. @fig-five-gau-projs a2, b2, c2 show the fitted models in a projection of the $4\text{-}D$ space. In this projection the difference between methods can be seen. These clusters are fully $4\text{-}D$ in nature, so we would expect the model to be a *crumpled \gD{} sheet* that stretches in all four dimensions. This is what is observed for tSNE and UMAP. The curious detail is that the model for PaCMAP is closer to a *pancake* in shape! What this means is that there has to be some ordering of points in the \gD{} PaCMAP layout that induces the flat model, likely that the points are organised by the global principal components. So the layout of the model in $4\text{-}D$ doesn't reflect the dimension of each cluster. DHC: is there something in the documentation that might help to explain what has happened?

>>>>

The models are fitted to these layouts, but we focus on a single cluster to illustrate the curious detail. @fig-five-gau-projs a2, b2, c2 show the fitted models in a projection of the $4\text{-}D$ space. In this projection the difference between methods can be seen. These clusters are fully $4\text{-}D$ in nature, so we would expect the model to be a *crumpled \gD{} sheet* that stretches in all four dimensions. This is what is observed for tSNE and UMAP. The curious detail is that the model for PaCMAP is closer to a *pancake* in shape! What this means is that there has to be some ordering of points in the \gD{} PaCMAP layout that induces the flat model, likely that the points are organised by the global principal components. So the layout of the model in $4\text{-}D$ doesn't reflect the dimension of each cluster. This can be happen because PaCMAP balances three components (near, mid-range, and far) rather than focusing solely on near neighbors. Mid-range neighbors ensure that transitions between clusters or groups are preserved and less concern with local preservation.

<!-- DHC: is there something in the documentation that might help to explain what has happened? -->
